---
title: "Estimation of sampling effort in community ecology with SSP"
author: "Edlin Guerra-Castro, Juan Carlos Cajas, Juan Jose Cruz-Motta, Nuno Simoes and Maite Mascaro"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimation of sampling effort in community ecology with SSP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}


knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`SSP` is an R package design to estimate sample effort in studies of ecological communities based on the definition of pseudo multivariate standard error (MultSE) (Anderson & Santana-Garcon 2015). This guide will provide you a brief overview in how to use `SSP`, as well as some description of the methodologies behind each function. The theoretical background are described in a submitted paper by Guerra-Castro et al. (2019).

The protocol in `SSP` consists in simulating several extensive data matrixes that mimic some of the relevant ecological features of the community of interest using a pilot data set. For each simulated data, several sampling efforts are repeatedly executed and MultSE calculated. The mean value, 0.025 and 0.975 quantiles of MultSE for each sampling effort across all simulated data are then estimated, and potentially ploted using `ggplot2`. The mean values are then standardized regarding the lowest sampling effort, and an optimal sampling effort can be identified as that in which the increase in sampling effort do not improve the precision beyond a threshold value (1, 2 or 3 %).
`SSP` include six functions: `assempar` for extrapolation of assemblage parameters using pilot data; `simdata` for simulation of several data sets based on extrapolated parameters; `datquality` for evaluation of plausibility of simulated data; `sampsd` for repeated estimations of MultSE for different sampling designs in simulated data sets; `summary_sd` for summarizing  the behaviour of MultSE for each sampling design across all simulated data sets, and `ioptimum` for identification of the optimal sampling effort.


## Software needed to run SSP in R
- Required: [R](https://cran.r-project.org/)
- Suggested: [RStudio IDE](https://www.rstudio.com/products/RStudio/#Desktop)

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))


